---
title: "bernes part"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```




```{r library, include=FALSE}
library(plyr)
library(tidyverse)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(reshape2)
library(stargazer)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(labelled)
library(readxl)
```



```{r q2 data, include=FALSE}

prescriber=read.csv('data\\prescriber-info.csv', stringsAsFactors=FALSE)
prescriber_list=c(prescriber$NPI)

#######2013##############
opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2013.csv', stringsAsFactors=FALSE)
opioids=subset(opioids, NPI %in% prescriber_list)
opioids=subset(opioids,Opioid.Claim.Count>1)
bernes<-opioids %>%
  dplyr::summarize(Opioids_sum= sum(`Opioid.Claim.Count`,na.rm=TRUE),
                   Total_claims = sum(`Total.Claim.Count`,na.rm=TRUE)) %>% 
  mutate(limit = Opioids_sum*0.8)%>%ungroup

opioids$limit<-bernes$limit
opioids<-arrange(opioids, desc(Opioid.Claim.Count))




opioids <-opioids%>%mutate(cum_claim = cumsum(`Opioid.Claim.Count`))
opioids$passed_limit<-0
opioids$passed_limit[opioids["cum_claim"]>opioids["limit"]]<-1


first_index<-min(which(opioids$passed_limit == 1))
pareto_2013=first_index/nrow(opioids)*100

top_10_opioids_prescriber_2013 <- opioids  %>% 
  select("NPI","Specialty.Description","NPPES.Provider.State","Opioid.Claim.Count")%>% 
  top_n(n = 20, wt = Opioid.Claim.Count)

names_2013<-c(top_10_opioids_prescriber_2013$NPI)


#######2014##############
opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2014.csv', stringsAsFactors=FALSE)
opioids=subset(opioids, NPI %in% prescriber_list)
opioids=subset(opioids,Opioid.Claim.Count>1)

bernes<-opioids %>%
  dplyr::summarize(Opioids_sum= sum(`Opioid.Claim.Count`,na.rm=TRUE),
                   Total_claims = sum(`Total.Claim.Count`,na.rm=TRUE)) %>% 
  mutate(limit = Opioids_sum*0.8)%>%ungroup

opioids$limit<-bernes$limit
opioids<-arrange(opioids, desc(Opioid.Claim.Count))



opioids <-opioids%>%mutate(cum_claim = cumsum(`Opioid.Claim.Count`))
opioids$passed_limit<-0
opioids$passed_limit[opioids["cum_claim"]>opioids["limit"]]<-1


first_index<-min(which(opioids$passed_limit == 1))
pareto_2014=first_index/nrow(opioids)*100

top_10_opioids_prescriber_2014 <- opioids  %>% 
  select("NPI","Specialty.Description","Opioid.Prescribing.Rate","Opioid.Claim.Count")%>%
  top_n(n = 20, wt = Opioid.Claim.Count)

names_2014<-c(top_10_opioids_prescriber_2014$NPI)


#######2015##############

opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2015.csv', stringsAsFactors=FALSE)
opioids=subset(opioids, NPI %in% prescriber_list)
opioids=subset(opioids,Opioid.Claim.Count>1)
bernes<-opioids %>%
  dplyr::summarize(Opioids_sum= sum(`Opioid.Claim.Count`,na.rm=TRUE),
                   Total_claims = sum(`Total.Claim.Count`,na.rm=TRUE)) %>% 
  mutate(limit = Opioids_sum*0.8)%>%ungroup

opioids$limit<-bernes$limit
opioids<-arrange(opioids, desc(Opioid.Claim.Count))



opioids <-opioids%>%mutate(cum_claim = cumsum(`Opioid.Claim.Count`))
opioids=subset(opioids, NPI %in% prescriber_list)
opioids$passed_limit<-0
opioids$passed_limit[opioids["cum_claim"]>opioids["limit"]]<-1


first_index<-min(which(opioids$passed_limit == 1))
pareto_2015=first_index/nrow(opioids)*100

top_10_opioids_prescriber_2015 <- opioids %>%
  select("NPI","Specialty.Description","Opioid.Prescribing.Rate","Opioid.Claim.Count" ) %>% 
  top_n(n = 20, wt = Opioid.Claim.Count)




names_2015<-c(top_10_opioids_prescriber_2015$NPI)

#######2016##############

opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2016.csv', stringsAsFactors=FALSE)
opioids=subset(opioids, NPI %in% prescriber_list)
opioids=subset(opioids,Opioid.Claim.Count>1)
bernes<-opioids %>%
  dplyr::summarize(Opioids_sum= sum(`Opioid.Claim.Count`,na.rm=TRUE),
                   Total_claims = sum(`Total.Claim.Count`,na.rm=TRUE)) %>% 
  mutate(limit = Opioids_sum*0.8)%>%ungroup

opioids$limit<-bernes$limit
opioids<-arrange(opioids, desc(Opioid.Claim.Count))



opioids <-opioids%>%mutate(cum_claim = cumsum(`Opioid.Claim.Count`))
opioids$passed_limit<-0
opioids$passed_limit[opioids["cum_claim"]>opioids["limit"]]<-1


first_index<-min(which(opioids$passed_limit == 1))
pareto_2016=first_index/nrow(opioids)*100

top_10_opioids_prescriber_2016 <- opioids %>% 
  select("NPI","Specialty.Description","Opioid.Prescribing.Rate","Opioid.Claim.Count")%>%
  top_n(n = 20, wt = Opioid.Claim.Count)

names_2016<-c(top_10_opioids_prescriber_2016$NPI)


top_dr<-Reduce(intersect, list(names_2013, names_2014,names_2015,names_2016))



######################tOP PRESCRIBERS###########################

opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2016.csv', stringsAsFactors=FALSE)

top_2016<-subset(opioids,NPI  %in% top_dr)
top_2016 <- top_2016 %>% 
  mutate(year = 2016) 

opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2015.csv', stringsAsFactors=FALSE)

top_2015<-subset(opioids,NPI  %in% top_dr)
top_2015 <- top_2015 %>% 
  mutate(year = 2015) 

opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2014.csv', stringsAsFactors=FALSE)

top_2014<-subset(opioids,NPI  %in% top_dr)
top_2014 <- top_2014 %>% 
  mutate(year = 2014) 

opioids=read.csv('data/Medicare_Part_D_Opioid_Prescriber_Summary_File_2013.csv', stringsAsFactors=FALSE)

top_2013<-subset(opioids,NPI  %in% top_dr)
top_2013 <- top_2013 %>% 
  mutate(year = 2013) 

cms_combined <- bind_rows(top_2013,top_2014,top_2015,top_2016)


```





```{r data model, include=FALSE}

model=read.csv('data/davis.csv', stringsAsFactors=FALSE)
gdp=read.csv('data/qgdpstate_all_R.csv', stringsAsFactors=FALSE)
#us_states=read.csv('data/us_states.csv', stringsAsFactors=FALSE)

gdp_long <- melt(gdp, id.vars = c("State", "Region", "ComponentName"))
gdp_long$year<-as.numeric(str_sub(gdp_long$variable, start= 2))
#long_combined <- merge(x=long, y=us_states, by.x='State',by.y='States', all.y=TRUE)

model_data <- merge(x=gdp_long, y=model, by.x=c('State',"year"),by.y=c('State',"year"), all.y=TRUE)

model_data<-model_data %>%
  mutate(gdp_percap=value*1000000/Population)





model_data<-model_data %>%
  mutate(opioid_cost_percap=opioid_cost/Population)



model_data<-model_data %>%
  mutate(male_female=male/female)




model_data<-model_data %>%
  mutate(white_other=white/(black+asian+hispanic+native_indian+other_race))


cms_combined_field<- cms_combined %>%filter(year==2013)%>%
  mutate(id=paste(NPPES.Provider.First.Name,NPPES.Provider.Last.Name,NPPES.Provider.State,Specialty.Description, sep="\n"))%>% 
  select(NPI,id)

cms_combined <- merge(x=cms_combined, y=cms_combined_field, by.x=c("NPI"),by.y=c("NPI"), all.y=TRUE)

```









### Q2 Pareto: top 20 percent of the prescribers provide 80% of the opioid in 2013


- Top **`r round(pareto_2013, digits = 2)`** percent of the prescribers provide 80% of the opioid in 2013.

- Top **`r round(pareto_2014, digits = 2)`** percent of the prescribers provide 80% of the opioid in 2014.

- Top **`r round(pareto_2015, digits = 2)`** percent of the prescribers provide 80% of the opioid in 2015.

- Top **`r round(pareto_2016, digits = 2)`** percent of the prescribers provide 80% of the opioid in 2016.


<div style="margin-bottom:100px;">
</div>


### Prescribers in top 20 for every years between 2013 to 2016

```{r top 20, echo = FALSE, fig.height = 21, fig.width=20,fig.align="center"}

ggplot(cms_combined, aes(x=year,
                         y=Opioid.Claim.Count)) + 
  facet_wrap(~ `id`) +
  geom_col() +
  xlab("Year") +
  ylab("Total Opioid claims") +
  theme(axis.text=element_text(size=18), axis.title=element_text(size=18),strip.text = element_text(size = 20))

```



```{r regression, echo = FALSE}

fit2 <- lm(Crude.Death.Rate ~ opioid_prescription_rate+average_age, data = model_data)

fit3 <- lm(Crude.Death.Rate ~ opioid_prescription_rate+
             white_other, data = model_data)

fit4 <- lm(Crude.Death.Rate  ~ opioid_prescription_rate+
             gdp_percap, data = model_data)

fit5 <- lm(Crude.Death.Rate ~ opioid_prescription_rate+average_age+white_other+
             gdp_percap, data = model_data)

fit6 <- lm(Crude.Death.Rate  ~ opioid_prescription_rate+average_age+white_other+
             gdp_percap+male_female, data = model_data)
```

### Regression Outputs
- Linear regression model use in all models.
- We could not utilize mixed or fixed effect regression since we only have four years for each state.
- Dependent variables is always crude death rate.

```{r results1, echo = FALSE}

sjt.lm(fit2,fit3,fit4,
       remove.spaces = TRUE,
       show.header = TRUE,
       show.ci = FALSE,
       p.numeric = FALSE, 
       show.se = T,
       digits.est = 3,
       CSS = list(css.separatorcol = 'padding-right:3em; padding-left:3em;'),
       string.est = "Estimate",
       string.ci = "Conf. Int.",
       string.p = "p-value",
       string.dv = "Response",
       string.pred = "Coefficients",
       string.interc = "Constant",
       group.pred = FALSE,
              pred.labels = c("Opioid Prescription Rate","Average Age", "White Other Race Ratio", "GDP per capita","Male Female Ratio"),
       depvar.labels = c("Crude Death Rate","Crude Death Rate","Crude Death Rate"))



```


<div style="margin-bottom:100px;">
</div>


```{r results2, echo = FALSE}

sjt.lm(fit5,fit6,
       remove.spaces = TRUE,
       show.header = TRUE,
       show.ci = FALSE,
       p.numeric = FALSE, 
       show.se = T,
       digits.est = 3,
       CSS = list(css.separatorcol = 'padding-right:3em; padding-left:3em;'),
       string.est = "Estimate",
       string.ci = "Conf. Int.",
       string.p = "p-value",
       string.dv = "Response",
       string.pred = "Coefficients",
       string.interc = "Constant",
       group.pred = FALSE,
              pred.labels = c("Opioid Prescription Rate","Average Age", "White Other Race Ratio", "GDP per capita","Male Female Ratio"),
       depvar.labels = c("Crude Death Rate","Crude Death Rate"))



```